---
title: "bst261 genomics example"
author: "Matt Ploenzke"
date: "5/7/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(keras)
library(ggseqlogo)
library(tidyverse)
library(stringi)
```

Model options.
```{r}
rm(list=ls(all=TRUE))
opt <- list()
opt$embedding1 <- 'CCCCCCC'
opt$max_len <- 200
opt$filter_len <- 16
```

Define some helper functions.
```{r}
one_hot <- function(dnastr) {
  a <- strsplit(as.character(dnastr),"")[[1]]
  buffer.left <- rep(0, ceiling((opt$max_len+2*opt$filter_len-2-nchar(dnastr))/2))
  buffer.right <- rep(0, floor((opt$max_len+2*opt$filter_len-2-nchar(dnastr))/2))
  oH <- c(buffer.left, as.numeric(a=="A"), buffer.right,
          buffer.left, as.numeric(a=="C"), buffer.right,
          buffer.left, as.numeric(a=="G"), buffer.right,
          buffer.left, as.numeric(a=="T"), buffer.right)
  return(oH)
}
```

Embed the sequence into positive and negative sequences respectively.
```{r}
negative.cases <- data.frame(sequence=sapply(1:1000,function(i) {paste(sample(c('A','C','G','T'),size=opt$max_len,replace=TRUE),collapse='')}),y=0)
positive.cases <- data.frame(sequence=sapply(1:1000,function(i) {paste(sample(c('A','C','G','T'),size=opt$max_len,replace=TRUE),collapse='')}),y=1)
positive.cases$sequence <- as.character(positive.cases$sequence)
for (irow in 1:nrow(positive.cases)) {
  location <- sample(1:(opt$max_len-nchar(opt$embedding1)),1)
  substr(positive.cases[irow,'sequence'],start=location,stop=(location+nchar(opt$embedding1)-1)) <- opt$embedding1
}
```

Join cases and randomly split into training and validation.
```{r}
all.cases <- rbind(positive.cases,negative.cases)
all.cases <- all.cases[sample(1:nrow(all.cases),size=nrow(all.cases),replace=FALSE),]
idx <- sample(1:nrow(all.cases),round(.9*nrow(all.cases)))
training_data <- all.cases[idx,c('sequence')]
training_labels <- all.cases[idx,'y']
testing_data <- all.cases[-idx,c('sequence')]
testing_labels <- all.cases[-idx,'y']
```

One-hot encode.
```{r}
training_array <- aperm(array(c(sapply(training_data,one_hot)),dim=c((opt$max_len+2*opt$filter_len-2),4,length(training_data),1)),perm=c(3,2,1,4))
testing_array <- aperm(array(c(sapply(testing_data,one_hot)),dim=c((opt$max_len+2*opt$filter_len-2),4,length(testing_data),1)),perm=c(3,2,1,4))
```

Define functions to output pretty plots during training.
```{r}
trackMotifs <- function(model, epoch) {
  all_dna_strings <- list()
  activations_fn <- k_function(inputs = list(model$input),outputs = list(model$get_layer(index=1L)$output))
  activations <- activations_fn(list(testing_array))[[1]]
  for (filt in 1:dim(activations)[4]) {
    act <- activations[,,,filt]
    act[act < .75*max(act)] <- 0 # this is the version used by Alipinahi
    pos.values <- which(act>0,arr.ind=TRUE)
    if (nrow(pos.values)>0) {
      seqs <- lapply(1:nrow(pos.values), function(seqq) {
        temp_str <- testing_data[pos.values[seqq,1]]
        temp_idx <- pos.values[seqq,2]
        temp_idx <- ifelse(temp_idx-opt$filter_len<1,temp_idx-(temp_idx-opt$filter_len),temp_idx)
        strseq <- substr(temp_str, temp_idx-(opt$filter_len-1), temp_idx)
        strseq <- ifelse(nchar(strseq)<opt$filter_len,paste(strseq,stri_dup("N",opt$filter_len-nchar(strseq)),sep=''),strseq)
        return(strseq)
      })
      dna_strings <- do.call(c,seqs)
      all_dna_strings[[paste('Motif',filt,sep=' ')]] <- dna_strings
    }
  }
  ggplot()+
    geom_logo(all_dna_strings,seq_type='dna') + 
    theme_logo() + 
    facet_wrap(~seq_group, ncol=ceiling(4), scales='free_x') + 
    theme(panel.background = element_rect(fill = NA, color = "black"),
          strip.text.x = element_text(size = 20),
          axis.title=element_text(size=20),
          axis.text=element_text(size=14),
          axis.text.x=element_text(size=10)) 
  ggsave("Empirical_filter_motifs.pdf",width=15,height=7.5,units='in')
}
```

R6 class to define callback.
```{r}
makeMotifs <- R6::R6Class("makeMotifs",
  inherit = KerasCallback,
  public = list(
    model = NULL,
    N = NULL,
    epoch = NULL,
    initialize = function(model = NA, N = NA) {
      self$model <- model
      self$N <- N
      self$epoch <- 1
    },
   on_epoch_end = function(epoch, logs = list()) {
     if (self$epoch %% self$N == 0) {
       trackMotifs(self$model, self$epoch)
     }
     self$epoch <- self$epoch + 1
   }
))
```

Build the model components including GC bias and full sequence information.
```{r}
full_sequence <- layer_input(shape=c(4,opt$max_len + 2*opt$filter_len-2,1))
full_sequence_model <- full_sequence %>%
  layer_conv_2d(filter = 8,
                kernel_size = c(4,opt$filter_len),
                padding = "valid",
                input_shape = c(4, opt$max_len+2*opt$filter_len-2, 1)) %>%
  layer_max_pooling_2d(pool_size = c(1,4)) %>%
  layer_flatten() %>% 
  layer_dense(units=1,activation='sigmoid') 

sequence_model <- keras_model(inputs = full_sequence, outputs = full_sequence_model)
sequence_model %>% compile(
  loss = "binary_crossentropy",
  optimizer = optimizer_adam(),
  metrics = c("accuracy")
)
```

Fit the model.
```{r}
moMotifs <- makeMotifs$new(sequence_model,1)
sequence_fit <- sequence_model %>% fit(
  x=training_array, training_labels,
  epochs = 10,
  validation_data = list(testing_array, testing_labels),
  shuffle = TRUE,
  callbacks = list(moMotifs)
)
```
